---
title: "Lecture 4 - Vectors and Matrices"
output: 
  xaringan::moon_reader:
    css: styles.css

date: "`r format(Sys.time(), '%d %B %Y')`"
author: "Ali Seyhun Saral <ali.saral@unibo.it>"
params:
    twitter: "@seyhunsaral"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---
```{r setup, include=FALSE}
# knitr::opts_chunk$set(collapse=TRUE, error=TRUE, warning=TRUE)
knitr::opts_chunk$set(dpi=300, warning = FALSE, fig.align = 'center', message = FALSE)
options(timeout= 4000000) 
```
## Recap of Last Week
* We've created vectors for plotting equations `x<- range(1,10)` and `y <- 3 * x ^ 2`

--

* We've learn how to create x-y plots using the built-in r functions `plot(x,y)`

--

* We've learned how to change plot parameters `plot(x,y, col="red", type="b")`

--

* We've learned how to create histograms `hist(x,y)`

--
* We've learned how to add lines on plots `abline(v = 10, col="red")`

---
# Vectors

$$\vec{a} = \left[\begin{array}{@{}c@{}}
    -1 \\
    1 \\
    \end{array} \right]$$

--

```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.6, every node/.style={transform shape}]
  \draw[thin,gray!40] (-2,-2) grid (2,2);
  \draw[<->] (-2,0)--(2,0) node[right]{$x$};
  \draw[<->] (0,-2)--(0,2) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\end{tikzpicture}
```

--

* Representing a vector in R

```{r}
a <- c(-1, 1)
```

---
## Sum of two vectors

```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.6, every node/.style={transform shape}]
  \draw[thin,gray!40] (-2,-2) grid (2,2);
  \draw[<->] (-2,0)--(2,0) node[right]{$x$};
  \draw[<->] (0,-2)--(0,2) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
  \draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
\end{tikzpicture}
```
---
## Sum of two vectors

```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.6, every node/.style={transform shape}]
  \draw[thin,gray!40] (-2,-2) grid (2,2);
  \draw[<->] (-2,0)--(2,0) node[right]{$x$};
  \draw[<->] (0,-2)--(0,2) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
  \draw[dashed, ->,>=stealth,red](-1,1)--(1,0) node[anchor=south west]{$\vec{\boldsymbol{b}}$};
\end{tikzpicture}
```

---
## Sum of two vectors

```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.6, every node/.style={transform shape}]
  \draw[thin,gray!40] (-2,-2) grid (2,2);
  \draw[<->] (-2,0)--(2,0) node[right]{$x$};
  \draw[<->] (0,-2)--(0,2) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
  \draw[dashed, ->,>=stealth,red](-1,1)--(1,0) node[anchor=south west]{$\vec{\boldsymbol{b}}$};
\draw[line width=1pt,green,-stealth](0,0)--(1,0) node[anchor=north east]{$\vec{\boldsymbol{a}}+\vec{\boldsymbol{b}}$};
\end{tikzpicture}
```
---
## Sum of two vectors

```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.4, every node/.style={transform shape}]
  \draw[thin,gray!40] (-2,-2) grid (2,2);
  \draw[<->] (-2,0)--(2,0) node[right]{$x$};
  \draw[<->] (0,-2)--(0,2) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
  \draw[dashed, ->,>=stealth,red](-1,1)--(1,0) node[anchor=south west]{$\vec{\boldsymbol{b}}$};
\draw[line width=1pt,green,-stealth](0,0)--(1,0) node[anchor=north east]{$\vec{\boldsymbol{a}}+\vec{\boldsymbol{b}}$};
\end{tikzpicture}
```

* Sum of two vectors in R

```{r}
a <- c(-1, 1)
b <- c(2, -1)
```

--

```{r}
c <- a + b
print(c)
```

---
## Subtraction of two vectors
* a - b
```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.3, every node/.style={transform shape}]
  \draw[thin,gray!40] (-3,-3) grid (3,3);
  \draw[<->] (-3,0)--(3,0) node[right]{$x$};
  \draw[<->] (0,-3)--(0,3) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
\end{tikzpicture}
```
---

## Subtraction of two vectors
* a - b
```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.3, every node/.style={transform shape}]
  \draw[thin,gray!40] (-3,-3) grid (3,3);
  \draw[<->] (-3,0)--(3,0) node[right]{$x$};
  \draw[<->] (0,-3)--(0,3) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
\draw[line width=1pt,orange,-stealth](0,0)--(-2,1) node[anchor=north east]{$\vec{\boldsymbol{-b}}$};
\end{tikzpicture}
```
---

## Subtraction of two vectors
* a - b
```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.3, every node/.style={transform shape}]
  \draw[thin,gray!40] (-3,-3) grid (3,3);
  \draw[<->] (-3,0)--(3,0) node[right]{$x$};
  \draw[<->] (0,-3)--(0,3) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
\draw[line width=1pt,orange,-stealth](0,0)--(-2,1) node[anchor=north east]{$\vec{\boldsymbol{-b}}$};
  \draw[dashed, ->,>=stealth,orange](-1,1)--(-3,2) node[anchor=south west]{$\vec{\boldsymbol{-b}}$};
\end{tikzpicture}
```
---


## Subtraction of two vectors
* a - b
```{r,engine='tikz',echo=FALSE}
\begin{tikzpicture}[thick,scale=0.3, every node/.style={transform shape}]
  \draw[thin,gray!40] (-3,-3) grid (3,3);
  \draw[<->] (-3,0)--(3,0) node[right]{$x$};
  \draw[<->] (0,-3)--(0,3) node[above]{$y$};
  \draw[line width=1pt,blue,-stealth](0,0)--(-1,1) node[anchor=south west]{$\vec{\boldsymbol{a}}$};
\draw[line width=1pt,red,-stealth](0,0)--(2,-1) node[anchor=north east]{$\vec{\boldsymbol{b}}$};
\draw[line width=1pt,orange,-stealth](0,0)--(-2,1) node[anchor=north east]{$\vec{\boldsymbol{-b}}$};
  \draw[dashed, ->,>=stealth,orange](-1,1)--(-3,2) node[anchor=south west]{$\vec{\boldsymbol{-b}}$};
\draw[line width=1pt,green,-stealth](0,0)--(-3,2) node[anchor=north east]{$\vec{\boldsymbol{a}}-\vec{\boldsymbol{b}}$};
\end{tikzpicture}
```
--

```{r}
a <- c(-1, 1)
b <- c(2, -1)
```

--

```{r}
c <- a - b 
print(c)
```
---
## Multiplying a Vector with a scalar
$$ \vec{b} = [b_1,b_2] $$
  $$ 3 \vec{b}= \[3b_1, 3b_2\]$$
--
  
```{r}
b <- c(1,2)
print(3 * b)
```
---
## Vector Multiplication
* The default multiplication in R <span style="color:red"> is not vector multiplication </span>

* It would multiply individual items separately.
--

```{r}
a <- c(2,3)
b <- c(1,2)
a * b
```
--

* The mathematical *vector/matrix multiplication* operator is: `%*%`.

```{r}
a %*% b
```
---
# Matrices 
\begin{pmatrix}
1 & 2 \\
3 & 4
\end{pmatrix}
---
```{r}
A <- matrix(c(1,2,3,4), nrow=2)
print(A)
```
--
* The default option lines values column-by-column
--
* You can add the option `byrow=TRUE` to give the values row-by-row
--
```{r}
A <- matrix(c(1,2,3,4), nrow=2, byrow=TRUE)
print(A)
```


