---
title: "Lecture 3 - Data Types, Vectors"
output: 
  xaringan::moon_reader:
    css: styles.css

date: "`r format(Sys.time(), '%d %B %Y')`"
author: "Ali Seyhun Saral <ali.saral@unibo.it>"
params:
    twitter: "@seyhunsaral"
---
```{r setup, include=FALSE}
# knitr::opts_chunk$set(collapse=TRUE, error=TRUE, warning=TRUE)
knitr::opts_chunk$set(dpi=300, warning = FALSE, fig.align = 'center', message = FALSE)
options(timeout= 4000000) 
```
## Recap of Last Week
* We've seen how to create vectors `c(item1,item2,item3)`

--

* We saw the data types we will use commonly: `logical` (`TRUE`,`FALSE`), `numeric` (`-21`,`42`,`32.5`), and `character` (`"its raining today"`)

--

* We've learned how to change one type to another: (`as.numeric("32")`, `as.character(32)`)

--

* We've seen two ways to create longer vectors: `seq(4,12, by=2)` and `rep(0,20)`.


---
## Basic Plots in R

Let's draw the points (1,1), (2,4), (3,8), (4,16), (5,12)

--


```{r}
x  <- c(1,2,3,4,5)
y  <- c(1,4,8,16,32)
```
--

```{r, eval=FALSE}
plot(x,y)
```

--
```{r echo=FALSE,fig.align = 'center', out.width = "40%"}
plot(x,y)
```

---

## Basic Plots in R

If we want to put a line between our observations

--

```{r}
x  <- c(1,2,3,4,5)
y  <- c(1,4,8,16,32)
```
--

```{r, eval=FALSE}
plot(x,y, type="l")

```

--
```{r echo=FALSE,fig.align = 'center', out.width = "40%"}
plot(x,y, type="l")
```
---

## We can draw both lines and points
```{r fig.align = 'center', out.width = "50%"}
plot(x,y, type="b")

```
---

## We can change shapes of the points
```{r fig.align = 'center', out.width = "50%"}
plot(x,y, type="b", pch = 24)

```
* pch: plotting character

---
## We can change sizes of the points
```{r fig.align = 'center', out.width = "50%"}
plot(x,y, type="b", pch = 24, cex=2)

```
* cex: character expansion

---

## We can change the color
```{r fig.align = 'center', out.width = "50%"}
plot(x,y, type="b", pch = 24, col="red", cex=2)

```

---

## Lets build the graph of $x^2$

1\. Create a vector with enough points, to be the x values (say from 1 to 100)

--

2\. Calculate the values for `x^2` and save it as another vector

--

3\. Plot those values as "line" 

---

## Lets build the graph of $x^2$

```{r, fig.align='center', out.width="50%"}
x  <- seq(0,100,0.1)
y  <- x ^ 2
plot(x,y, type="l")
```

---

## Histograms

```{r, eval=FALSE}

heights  <- c(1.75, 1.80, 1.82,1.55,1.84,1.85 ) # and more

```

```{r, echo=FALSE}
heights <- rnorm(50, 1.75, 0.10)
```

--

```{r, fig.align='center', out.width="50%"}
hist(heights)
```

---

## Histograms 

Let's calculate the average height

--

```{r}
mean_heights <- mean(heights)
```

--

```{r}
print(mean_heights)
```

---

## Vertical line on the plot

```{r, fig.align='center', out.width="50%"}
hist(heights)
abline(v = mean_heights, col="red")

```

---

## Vertical line on the plot
```
abline(v= vertical_value, h=horizontal_value)
```


* `abline()` places on top of the histogram
* It is common in R to use `plot(x,y)` and then place other layers on top


```{r, fig.align='center', out.width="40%"}
hist(heights)
abline(v = mean_heights, col="red")

```
    
---

## Classroom Experiment

- How many candies do you think there are in the jar?
<center>
<h1> www.saral.it/jar </h1>
</center>

---

## Excercises

1 - Put the data of the guesses in a vector. (if you can read from csv file... good. but a simple copy past will suffice)


2 - Calculate the mean and save it to `mean_guesses`


3 - Draw a histogram from the guesses data


4 - Put the real number as a vertical line on the plot.

---

5 - Think about the results. Do you think it was accurate? Were you expecting it? How well did you do?

6 - Create the vector of deviations (differences) from the real number. Save it as `accuracies`

7 - Now create a vector from the seating row data. save it as `seatings`

8 - Plot a graph with `seatings` on the x axis and `accuracies` on the y axis.

9 - Interpret the graph. Do you think the people on the front rows did better?
