---
title: "Lecture 5 - Matrices (con'td)"
output: 
  xaringan::moon_reader:
    css: styles.css

date: "`r format(Sys.time(), '%d %B %Y')`"
author: "Ali Seyhun Saral <ali.saral@unibo.it>"
params:
    twitter: "@seyhunsaral"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---
```{r setup, include=FALSE}
# knitr::opts_chunk$set(collapse=TRUE, error=TRUE, warning=TRUE)
knitr::opts_chunk$set(dpi=300, warning = FALSE, fig.align = 'center', message = FALSE)
options(timeout= 4000000) 
```
## Recap of Last Week
* We've seen some vector operations.
--

* We've learned how to create matrices from a vector: `matrix(c(1,2,3,4), nrow=2)`

--

* We've learned how to create special matrices:
    * Same elements: `matrix(element, nrow, ncol)`
    * Identity matrix: `diag(n)`

--
* We've learned how to multiply matrices  `%*%`


--
* We've learned how to find the inverse of a matrix `solve(M)`


--

---
# Matrix Determinants

$$
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}
$$


* The formula for the determinant for the 2x2:

$$|A| = ad - bc $$
---

* For the larger matrices, we can break down matrices into 2x2 matrices and find the determinant. 

$$
\begin{bmatrix}
a & b & c\\
d & e & f \\
g & h & i
\end{bmatrix}  = a \begin{vmatrix}
e & f \\
h  & i
\end{vmatrix} - b \begin{vmatrix}
d & f \\
g  & i
\end{vmatrix} +  c \begin{vmatrix}
d & e \\
g  & h
\end{vmatrix}
$$
---
# Determinants

* $Det(A) = 0$ : Singular (not invertible)

--

* $Det(A) \neq 0$: Invertible

---
## Determinants
* In R, you can use `det()` function to find the determinant
```{r}
B = matrix(c(-1,2,4,3), byrow = TRUE, nrow= 2)
print(B)

det(B)

```
---

## Properties of a unique solution
$$ A\vec{x} = \vec{b} $$

- Matrix A has inverse (is invertible)
- Determinant of A is nonzero
- The rows and columns of A form a basis for the set of all vectors with n elements

```{r, error=TRUE}
A  <- matrix(c(1,2,4,-1,1,-2), nrow=3)
A
solve(A)
```
---

```{r, error=TRUE}
B  <- matrix(c(1,2,0,0), nrow=2)
B
det(B)
solve(B)
```
---

```{r, error=TRUE}
C <- matrix(c(1,2,1,0), nrow=2)
C
solve(C)
```
---

$$ C\vec{x} = \vec{b} $$
$$ C = 
\begin{pmatrix}
1 & 1 \\
2 & 0
\end{pmatrix}  $$

--

$$ b = \begin{bmatrix}
1 \\
2 
\end{bmatrix}  $$
$$ x = ? $$
--

```{r}
x  <- matrix(c(1,2),nrow=2)
C  <- matrix(c(1,2,1,0), nrow=2)

b  <- solve(C)  %*% x
b

# Let's double check
C  %*% b
```
---
# Excercise 

* Consider the system:

$$
9x + 10y = 34  \\ 
$$

$$
-6x - 5y = -26
$$

* Write the system in a matrix form using R in the form Ax = B. (Don't write the vector of unknowns).

* Check if the system has a solution by
 * Matrix has A has inverse
 * Determinant A is non-zero